 

### 2023-07-31 markdown/article.Rmd
article first draft finished

With the growing prominence of single-cell techniques across various omics fields, there is a pressing need to develop a standardized pipeline for proteomics data in the realm of systems biology. Unlike DNA sequencing and RNA sequencing, proteomics analysis via mass spectrometry incurs high costs in terms of labor and equipment. Additionally, commercially available software solutions often come with hefty price tags and limited transparency regarding the underlying methods employed. However, MaxQuant \citep{Cox2008} presents a promising alternative, complemented by the flexibility using the R programming language. 
Introducing ProteoScanR, a state-of-the-art proteomics pipeline integrated into the user-friendly Proteomics Workbench interface. 
Guided by SCoPE2 \citep{Specht2021, Petelski2021, Vanderaa2021}, the development process of ProteoScanR was thoroughly tested and validated using both bulk and single-cell mass spectrometry data sets. The pipeline implementation in the Proteomics Workbench leverages the power of an interactive environment built in R Shiny, empowering users to discover valuable insights for their specific data set. 
Within the interactive environment, users have the flexibility to customize cutoffs and thresholds for quality control, as well as employ various approaches for data transformation, normalization, missing value imputation, and batch correction. 
ProteoScanR and the Proteomics Workbench serves as a valuable tool in guiding the identification of expressed proteins in cells under study. Subsequently, the pathway enrichment analysis provides additional biological contexts for a comprehensive understanding of their functional implications.
The master thesis project serves as a foundation for future advancements in the field of single-cell proteomics. Moreover, the codebase has been designed with robustness and scalability in mind, ensuring ease of maintenance and future expansion of the application
The source code is accessible and can be located at the following URL: https://github.com/Lukas67/ProteoScanR

@@

Preparation for hypothesis testing involves data transformation and normalization. Missing values for proteins in individual channels are handled according to user selection with methods such as the K-nearest neighbor algorithm (=KNN) \citep{Lan2013}. Before statistical analysis, dimensionality reduction techniques can be applied and observed with or without correcting for batch effect by ComBat \citep{Johnson2007}. Furthermore data processing can be validated with an entropy based approach for the conservation of information.

@@

After finding differentially expressed proteins, enrichment elucidates them in biological context by finding over represented proteins. The protein enrichment analysis can be employed in selected contexts such as pathways, cell-types or custom ontologies. The pathway enrichment analysis is performed using the R Bioconductor package clusterProfiler \citep{Wu2021}. The analysis utilizes the entire “Kyoto Encyclopedia of Genes and Genomes” (KEGG) database, which is accessed through a function that maps the UniProt IDs to corresponding pathways. For custom ontologies the protein enrichment analysis is done using the R bioconductor package piano \citep{Vaeremo2013}. In opposite to the pathway enrichment, piano maps proteins to a pre-selected gene collection set, which can be either downloaded from gsea-msigdb.org or custom made for the particular experiment and explains conditions, phenotype or other desired properties.

@@

As seen in figure \@ref(fig:ui-numbers-default) the aspects shown in the main panel include:
\begin{scriptsize}
\begin{itemize}
\item Overview showing the individual processing steps of the analysis pipeline
\item Barchart showing the number of peptide spectrum matches, peptides and proteins
\item Mutual information intra- and inter- groupwise
\item Reporter ion intensity with factor selection to highlight
\item Median coefficient of variation for razor proteins and correlation heatmap for protein expression
\item Dimensionality reduction analysis with principle component analysis (PCA) and Uniform Manifold Approximation and Projection (UMAP)
\item Feature wise output for an individual protein over the course of the pipeline.
\item Statistics module
\item Pathway enrichment
\item Protein set ontology enrichment 
\end{itemize}
\end{scriptsize}

>>>>

With the growing prominence of single-cell techniques across various omics fields, there is a pressing need to develop a standardized pipeline for proteomics data in the realm of systems biology. Unlike DNA sequencing and RNA sequencing, proteomics analysis via mass spectrometry incurs high costs in terms of labor and equipment. Additionally, commercially available software solutions often come with hefty price tags and limited transparency regarding the underlying methods employed. However, MaxQuant \citep{Cox2008} presents a promising alternative, complemented by the flexibility using the R programming language. 
Introducing ProteoScanR, a state-of-the-art proteomics pipeline integrated into the user-friendly Proteomics Workbench interface. 
Guided by SCoPE2 \citep{Specht2021, Petelski2021, Vanderaa2021}, the development process of ProteoScanR was thoroughly tested and validated using both bulk and single-cell mass spectrometry data sets. The pipeline implementation in the Proteomics Workbench leverages the power of an interactive environment built in R Shiny, empowering users to discover valuable insights for their specific data set. 
Within the interactive environment, users have the flexibility to customize cutoffs and thresholds for quality control, as well as employ various approaches for data transformation, normalization, missing value imputation, and batch correction. 
ProteoScanR and the Proteomics Workbench serves as a valuable tool in guiding the identification of expressed proteins in cells under study. Subsequently, the pathway enrichment analysis provides additional biological contexts for a comprehensive understanding of their functional implications.
The master thesis project serves as a foundation for future advancements in the field of single-cell proteomics. Moreover, the codebase has been designed with robustness and scalability in mind, ensuring ease of maintenance and future expansion of the application.
The source code is accessible and can be located at the following URL: https://github.com/Lukas67/ProteoScanR

@@

Preparation for hypothesis testing involves data transformation and normalization. Missing values for proteins in individual channels are handled according to user selection with methods such as the K-nearest neighbor algorithm (=KNN) \citep{Lan2013}. Before statistical analysis, dimensionality reduction techniques were applied and observed with and without correcting for batch effect by ComBat \citep{Johnson2007}. Furthermore data processing is validated with an entropy based approach for the conservation of information.

@@

After finding differentially expressed proteins, enrichment elucidates them in biological context by finding over represented proteins. The protein enrichment analysis is employed in selected contexts such as pathways, cell-types or custom ontologies. The pathway enrichment analysis is performed using the R Bioconductor package clusterProfiler \citep{Wu2021}. The analysis utilizes the entire “Kyoto Encyclopedia of Genes and Genomes” (KEGG) database, which is accessed through a function that maps the UniProt IDs to corresponding pathways. For custom ontologies the protein enrichment analysis undertaken with the R bioconductor package piano \citep{Vaeremo2013}. In opposite to the pathway enrichment, piano maps proteins to a pre-selected gene collection set, which is either downloaded from gsea-msigdb.org or custom made for the particular experiment and explains conditions, phenotype or other desired properties.

@@

As seen in figure \@ref(fig:ui-numbers-default) the aspects shown in the main panel include:
\spacing{1}
\begin{scriptsize}
\begin{itemize}
\item \textbf{Overview:} Summary showing the individual processing steps of the analysis pipeline
\item \textbf{Summary Barplot:} Barchart showing the number of peptide spectrum matches, peptides and proteins
\item \textbf{Normalization validation:} Mutual information intra- and inter- groupwise
\item \textbf{Reporter ion intensity:} Includes factor selection to create boxplots indicating a factor
\item \textbf{Median coefficient of variation and correlation:} Variation of peptides for razor proteins and correlation heatmap for protein expression
\item \textbf{Dimensionality reduction:} Analysis with principle component analysis (PCA) and Uniform Manifold Approximation and Projection (UMAP)
\item \textbf{Feature wise output:} Visualization of an individual protein over the course of the pipeline.
\item \textbf{Statistics module:} Differential expression analysis
\item \textbf{Pathway enrichment:} Over-representation analysis for differentially expressed proteins in the context of pathways.
\item \textbf{Protein set ontology enrichment:}  Over-representation analysis for differentially expressed proteins in a user selected context.
\end{itemize}
\end{scriptsize}
\spacing{1.5}
